How to Not write Virtual Machines for Dynamic Languages

### 2007-05-27 talk/dyla2007/dyla.tex
Remove one XXX, add a (less purely aesthetic) one. Progress!
just remove one XXX. we have space problems anyway.
Fix refs.

In section \ref{sect:approaches} we will explore the way VMs are typically
implemented in C and on top of OO VMs and some of the problems of these
approaches, using various Python implementations as the main example. In
\ref{section sect:metaprogramming} we will describe our proposed
meta-programming approach.

\section{Approaches to Dynamic Language Implementation}
\label{sect:appraches}

@@

XXX mention that OO languages don't have good metaprogramming support

@@

\item
\emph{Ease of implementation:} This point is disputable. On the one hand, OO
VMs typically allow the language implementor to start at a higher level. On the
other hand they also enforce a specific object and execution model. This means
that the concepts of the implemented language need to be mapped to the
execution model of the underlying VM, which may be easy or not, depending very
much on the language in question. XXX give example where it works well.

An example where this mapping does not work too well is Prolog. While there
exist several implementations of Prolog on top of the JVM \cite{prologcafe}
\cite{InterProlog} and also one on .NET \cite{psharp},
they are not particular efficient, especially when compared to good Prolog VMs
in written in C. This is mostly because the Prolog execution model, which
involves backtracking and deep recursion does not fit the JVM and .NET very
well. Therefore the Prolog implementations on top of OO VMs resort to models
that is quite unnatural both for the OO VM and for Prolog.
\end{itemize}

@@

\begin{itemize}
\item \emph{Do not write dynamic language implementations ``by hand''.}
Writing them more abstractly, at a higher level, has mostly only
advantages, among them avoiding a proliferation of implementations
growing out of sync.  Writing interpreters both flexibly and efficiently
is difficult, and meta-programming is a good way to achieve it.
Moreover, this is not incompatible with targetting and benefiting from
existing good object-oriented virtual machines like the Java and .NET
ones.

>>>>

In section \ref{sect:approaches} we will explore the way VMs are typically
implemented in C and on top of OO VMs and some of the problems of these
approaches, using various Python implementations as the main example. In
section \ref{sect:metaprogramming} we will describe our proposed
meta-programming approach.

\section{Approaches to Dynamic Language Implementation}
\label{sect:approaches}

@@

@@

\item
\emph{Ease of implementation:} This point is disputable. On the one hand, OO
VMs typically allow the language implementor to start at a higher level. On the
other hand they also enforce a specific object and execution model. This means
that the concepts of the implemented language need to be mapped to the
execution model of the underlying VM, which may be easy or not, depending very
much on the language in question.

An example where this mapping does not work too well is Prolog. While there
exist several implementations of Prolog on top of the JVM \cite{prologcafe}
\cite{InterProlog} and also one on .NET \cite{psharp},
they are not particular efficient, especially when compared to good Prolog VMs
in written in C. This is mostly because the Prolog execution model, which
involves backtracking and deep recursion does not fit the JVM and .NET very
well. Therefore the Prolog implementations on top of OO VMs resort to models
that is quite unnatural both for the OO VM and for Prolog.

Another important point that make implementations of languages on top of OO VMs
harder is that typically they don't support meta-programming very well, or only
at the bytecode level.
\end{itemize}

@@

%XXX doesn't look entirely nice
\begin{itemize}
\item \emph{Do not write dynamic language implementations ``by hand''.}
Writing them more abstractly, at a higher level, has mostly only
advantages, among them avoiding a proliferation of implementations
growing out of sync.  Writing interpreters both flexibly and efficiently
is difficult, and meta-programming is a good way to achieve it.
Moreover, this is not incompatible with targetting and benefiting from
existing good object-oriented virtual machines like the Java and .NET
ones.
